---
title: "trainR: basic example explained"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{trainR: basic example explained}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed( 000 )
```

# Overview of `trainR`

trainR is a package that helps the user organize and implement efficient,
augmented training with ANTsRNet.  the primary target audience is those who
are working with volumetric data but 2D will work too.  the basic idea is
to make training more efficient through numpy-based I/O of the "infinite"
training data that we can often generate.

the augmentation strategy will use:

* geometric transformation of images (rigid, affine, etc)

* gaussian noise

* simulated bias field

* histogram warping

to create new versions of images from an input image list.
Segmentations will be augmented as well and there are options for treating
the segmentations as points.  One can convert points (in physical space)
to segmentation images via `ANTsR` function `makePointsImage` or by converting
physical points to an index via `antsTransformPhysicalPointToIndex`.  If one
uses the latter approach, one should encode each point by a unique label.

## Setting up the data

**Run on both threads when setting up**

Environment variables control multi-threading for CPU in ANTs and tensorflow.
See below for how to set these at the beginning on an experiment.  We set
the number of threads to 12 below.

```{r setup2,echo=TRUE,eval=FALSE}
Sys.setenv("TF_NUM_INTEROP_THREADS"=12)
Sys.setenv("TF_NUM_INTRAOP_THREADS"=12)
Sys.setenv("ITK_GLOBAL_DEFAULT_NUMBER_OF_THREADS"=12)
```


See the code below for how to assemble the images for augmentation.

**Run on both threads when setting up**

```{r setupX,echo=TRUE,eval=FALSE}
# record the GPU
gpuid = Sys.getenv(x = "CUDA_VISIBLE_DEVICES")
library( ANTsR )
library( ANTsRNet )
library( tensorflow )
library( keras )
library( tfdatasets )
library( reticulate )
library( patchMatchR )
library( trainR )
np <- import("numpy")
mytype = "float32"
nlayers = 4   # for unet
downsam = 4   # downsamples images
# collect your images - a list of lists - multiple entries in each list are fine
rids <- function( x, downsammer=4 ) {
  img = ri( x )
  resampleImage( img, dim(img)/downsammer, useVoxels=TRUE )
}
upperhalf = rids( 1 ) * 0 + 1 # this just gives us more segmentation labels
upperhalf[33:64,1:32] = 2
upperhalf[1:32,33:64] = 3
upperhalf[33:64,33:64] = 4
ilist = list(
  list( rids( 1 ), getMask( rids( 1 ) ) ),
  list( rids( 2 ), getMask( rids( 2 ) ) ),
  list( rids( 3 ), getMask( rids( 3 ) ) ),
  list( rids( 4 ), getMask( rids( 4 ) ) ),
  list( rids( 6 ), getMask( rids( 6 ) ) ) )

slist = list()
for ( k in 1:length( ilist ) ) {
  temp = thresholdImage( ilist[[k]][[1]], "Otsu", 3 )
  temp = temp * ( ilist[[k]][[2]] - iMath( ilist[[k]][[2]], "ME", 6 ) )
  slist[[k]] = thresholdImage( temp, 2, 2 ) * upperhalf
  plot( ilist[[k]][[1]], slist[[k]] )
}
```

Here are definitions for some variables important for training: the numbers
defining the segmentation labels and which images are for training and testing.

**Run on both threads when setting up**

```{r setup3,echo=TRUE,eval=FALSE}
# identify the number of segmentation classes
segregions = sort( unique( slist[[1]]  ))[-1] # the seg numbers
isTrain = c( rep(TRUE,length(slist)-1), FALSE )
```

This example will do landmark learning.  If doing segmentation, then include
the zero label i.e. just `sort( unique( slist[[1]]  ))`.


## Preparing to write the data to disk

We will generate (in this case) 5 files.  For each, we generate a random file
unique ID and then create file extensions that map to the type of augmented
data we will be generating.  In this case, we will generate augmented images,
the associated masks, the points, coord conv output and a heatmap.  These
all derive from the image lists we defined above.  This stuff will be run
on CPU separately from the training loop.

**Run on CPU thread when setting up**

```{r disk,echo=TRUE,eval=FALSE}
nFiles = 5
if ( ! exists( "uid" ) )
  uid = paste0("LM",sample(1:1000000,nFiles),"Z")
types = c("Images.npy", "Points.npy", "mask.npy", "coordconv.npy", "heatmap.npy" )
nmats = length(types) # 5 arrays out
trainTestFileNames = data.frame( matrix("",nrow=nFiles,ncol=nmats*2))
colnamesTT = c(
  paste0("train",gsub(".npy","",types)),
  paste0("test",gsub(".npy","",types)) )
colnames( trainTestFileNames ) = colnamesTT
for ( k in 1:nFiles ) {
  twogroups = paste0("numpy/",uid[k],c("train","test"))
  npextsTr = paste0( twogroups[1], types )
  npextsTe = paste0( twogroups[2], types )
  trainTestFileNames[k,]=as.character( c(npextsTr,npextsTe) )
}
write.csv( trainTestFileNames, "numpy/LMtrainttestfiles.csv", row.names=FALSE)
```


## Augmenting the data

We generate the data based on default parameters. If you want to learn more about
the possibilities here, then please change the parameters and take a look at the
results.   We will show an example that looks at some parameter variations below.

The call below will write a test file to numpy.  We just write one in this
case as we keep it constant during training.  Write as many as you want for
your application

**Run on CPU thread when setting up**

```{r augment,echo=TRUE,eval=FALSE}
testfilename = as.character(trainTestFileNames[1,grep("test",colnames(trainTestFileNames))])
gg = generateDiskData(
    inputImageList = ilist,
    segmentationImageList = slist,
    segmentationNumbers = segregions,
    selector = !isTrain,
    addCoordConv = TRUE,
    segmentationsArePoints = TRUE,
    smoothHeatMaps = 3.0,
    maskIndex = 2,
    transformType = "scaleShear",
    noiseParameters = c(0, 0.05),
    sdSimulatedBiasField = 0.01,
    sdHistogramWarping = 0.01,
    sdAffine = 0.1, # limited
    numpynames = testfilename,
    numberOfSimulations = 8
    )
# visualize example augmented images
layout( matrix(1:8,nrow=2))
for ( k in 1:8 ) {
  temp = as.antsImage( gg[[1]][k,,,1] )
  plot( temp )
  }
```


We would typically run something like the code below in a continuous loop on
a CPU thread separate from the GPU training loop.  In practice, you may keep
the parameters the same across train and test or set minimal augmentation
parameters for test data.

**Run on CPU thread when setting up**

```{r augmenttrain,echo=TRUE,eval=FALSE}
while( TRUE ) {
  for ( k in 1:nFiles ) {
    trnfilename = as.character(trainTestFileNames[k,grep("train",colnames(trainTestFileNames))])
    print( paste(k, trnfilename[1] ) )
    gg = generateDiskData(
        inputImageList = ilist,
        segmentationImageList = slist,
        segmentationNumbers = segregions,
        selector = isTrain,
        addCoordConv = TRUE,
        segmentationsArePoints = TRUE,
        smoothHeatMaps = 3.0,
        maskIndex = 2,
        transformType = "scaleShear",
        noiseParameters = c(0, 0.05),
        sdSimulatedBiasField = 0.01,
        sdHistogramWarping = 0.01,
        sdAffine = 0.1,
        numpynames = trnfilename,
        numberOfSimulations = 64
        )
    }
  }
```

Getting the above steps correct is very critical to successful training.
This package currently hides some of that complexity for basic segmentation and
landmark prediction.  However, it is always possible to make mistakes if the
training decisions are inconsistent with the underlying theory.  A forthcoming
publication will explain the theory associated with this package.

## set up the unet

Below is the default set up for unet-based lanmdark prediction given
the above augmentation choices.

**Run on GPU thread when setting up**

```{r unetsetup,echo=TRUE,eval=FALSE}
library( ANTsR )
library( ANTsRNet )
library( tensorflow )
library( keras )
library( tfdatasets )
library( reticulate )
library( patchMatchR )
library( trainR )
np <- import("numpy")
mytype = "float32"
nlayers = 4   # for unet
downsam = 4   # downsamples images
# set up the network - all parameters below could be optimized for the application
unet = createUnetModel2D(
       list( NULL, NULL, 1 ),
       numberOfOutputs = length( segregions ),
       numberOfLayers = nlayers, # should optimize this wrt criterion
       numberOfFiltersAtBaseLayer = 32, # should optimize this wrt criterion
       convolutionKernelSize = 3, # maybe should optimize this wrt criterion
       deconvolutionKernelSize = 2,
       poolSize = 2,
       strides = 2,
       dropoutRate = 0,
       weightDecay = 0,
       additionalOptions = c( "nnUnetActivationStyle" ),
       mode = c("regression")
     )
maskinput = layer_input( list( NULL, NULL, 1 ) )
posteriorMask <- layer_multiply(
  list( unet$outputs[[1]] , maskinput ), name='maskTimesPosteriors'  )
unet = keras_model( list( unet$inputs[[1]], maskinput ), posteriorMask )
unetLM = patchMatchR::deepLandmarkRegressionWithHeatmaps( unet,
  activation = 'none', theta=0 )
unetLM1 = patchMatchR::deepLandmarkRegressionWithHeatmaps( unet,
  activation = 'relu', theta=0 )
```

## Reading the data from disk

Now we have to read the data from disk in a parallel thread.  We handle this
by looking at the file modification times and we, by default, choose to read
the file that was modified *2nd* most recently.

The functions below handle how to choose which image to read for training.
Some experimentation may be needed to get the timing correct.  You want to
balance the time it takes to augment versus the time it takes to train
through a batch.

**Run on GPU thread when setting up**

```{r howdowereadintime,echo=TRUE,eval=FALSE}
trtefns = read.csv( "numpy/LMtrainttestfiles.csv" ) # critical - same file name
trnnames = colnames(trtefns)[grep("train", colnames(trtefns) )]
tstnames = colnames(trtefns)[grep("test", colnames(trtefns) )]
loadfirst = chooseTrainingFilesToRead( trtefns[,trnnames] )
Xtr = trainR::loadNPData( loadfirst )
mybs = dim( Xtr[[1]] )[1]
Xte = trainR::loadNPData( trtefns[1,tstnames] )
```

spatial tensors are indexed in the following way:

* first index: subject (or batch) index
* second index: x coordinate
* third index : y coordinate
* fourth index : z or channels
* last/fifth index (if available): channels

e.g in this example we see:

```
> dim( Xtr[[1]] )
[1] 64 64 64  1
```

which is the size of the batch (64) and then spatial dimension of the input image (64x64) with just one channel.

point coordinate tensors are indexed in the following way:

* first index: subject (or batch) index
* second index: which landmark we are going to index (eg the first, second etc)
* third index : the point / landmark coordinate itself which will be a 2 vector in 2D and 3 vector in 3D

e.g.
```
> dim( Xtr[[2]] )
[1] 64  4  2
```
which is the size of the batch (64) and then number of landmarks (4 here) with 2 spatial dimensions.


## set up the training history data frame

**Run on GPU thread when setting up**

```{r seghistory,echo=TRUE,eval=FALSE}
mydf = data.frame()
epoch = 1
wtfn=paste0('lm_weights_gpu', gpuid,'.h5')
csvfn = paste0('lm_weights_gpu', gpuid,'.csv')
```

## Using the data in a training loop for a neural network

below is a "standard" tensorflow training loop with some alterations
that make some efficiencies / choices a little more explicit but still clear.
at least, that is what's intended.

**Run on GPU thread when setting up**

```{r training,echo=TRUE,eval=FALSE}
epoch = 1
for ( ptwt in c( 0.001, 0.005, 0.01, 0.05, 0.1 ) ) {
  if ( ptwt == 0.01 ) unetLM = unetLM1
  ptWeight = tf$cast( ptwt, mytype )
  num_epochs <- 100
  if ( ptwt == 0.1 ) num_epochs = 1500
  optimizerE <- tf$keras$optimizers$Adam(1.e-6)
  batchsize = 2
  for (epoch in 1:num_epochs ) {
    if ( (epoch %% round(mybs/batchsize) ) == 0 & epoch > 1  ) {
        # refresh the data
        locfns = chooseTrainingFilesToRead( trtefns[,trnnames] )
        print( locfns[1] )
        Xtr = trainR::loadNPData( locfns )
      }
    ct = nrow( mydf ) + 1
    mysam = sample( 1:nrow(Xtr[[1]]), batchsize )
    datalist = list()
    datalist[[2]] = array( Xtr[[2]][mysam,,], dim=dim(Xtr[[2]][mysam,,]) ) %>% tf$cast( mytype )
    for ( jj in c(1,3:5) )
      datalist[[jj]] = array( Xtr[[jj]][mysam,,,],
        dim=c(batchsize,tail(dim(Xtr[[jj]]),3)) ) %>% tf$cast( mytype )
    with(tf$GradientTape(persistent = FALSE) %as% tape, {
      preds = unetLM( datalist[c(1,3:4)] )
      lossht = tf$keras$losses$mse( datalist[[5]], preds[[1]] ) %>% tf$reduce_mean( )
      losspt = tf$keras$losses$mse( datalist[[2]], preds[[2]] ) %>% tf$reduce_mean( )
      loss = losspt * ptWeight + lossht
      })
    unet_gradients <- tape$gradient(loss, unetLM$trainable_variables)
    optimizerE$apply_gradients(purrr::transpose(list(
        unet_gradients, unetLM$trainable_variables )))
    mydf[ct,'train_loss'] = as.numeric( loss )
    mydf[ct,'train_ptloss'] = as.numeric( losspt )
    mydf[ct,'train_htloss'] = as.numeric( lossht )
    mydf[ct,'ptWeight'] = as.numeric( ptWeight )
    if( epoch > 3 & epoch %% 10 == 0 ) {
      with(tf$device("/cpu:0"), {
        preds = predict( unetLM, Xte[c(1,3:4)] )
        lossht = tf$keras$losses$mse( Xte[[5]], preds[[1]] ) %>% tf$reduce_mean( )
        losspt = tf$keras$losses$mse( Xte[[2]], preds[[2]] ) %>% tf$reduce_mean( )
        loss = tf$cast(losspt, mytype) * ptWeight + tf$cast(lossht, mytype)
      })
      # compute the same thing in test data
      mydf[ct,'test_loss'] = as.numeric( loss )
      mydf[ct,'test_ptloss'] = as.numeric( losspt )
      mydf[ct,'test_htloss'] = as.numeric( lossht )
      if ( mydf[ct,'test_ptloss'] <= min(mydf[1:epoch,'test_ptloss'],na.rm=TRUE) ) {
        print(paste("Saving",epoch))
        keras::save_model_weights_hdf5( unetLM, wtfn )
        gc()
      }
    }
  print( mydf[ct,] )
  write.csv( mydf, csvfn, row.names=FALSE )
  }
}
```

## monitoring training progess

take a look at the training-test convergence curves.

**Run on CPU thread**

```{r traintestcurves,eval=FALSE,echo=TRUE}
if ( ! exists( "mydf" ) ) mydf = read.csv( 'lm_weights_gpu.csv' )
mydfnona = mydf[ !is.na( mydf$test_loss),  ]
plot( ts( mydfnona ) )
```



## show the predicted points in the test data

Visualize the predicted points.

```{r vizp,echo=TRUE,eval=FALSE}
layout(matrix(1:2,nrow=1))
wsub = 1
testimg = as.antsImage( Xte[[1]][wsub,,,1] ) %>%
  antsCopyImageInfo2( rids( 1 ) )
preds = predict( unetLM, Xte[c(1,3:4)] )
# these are in physical space
predPoints = as.array( preds[[2]] )[wsub,,]
truPoints = Xte[[2]][wsub,,] # true points
print(paste("Error",norm(truPoints-predPoints,"F")))
print(paste("%Error",norm(truPoints-predPoints,"F")/norm(truPoints,"F")*100,"%"))
truIndex = round( antsTransformPhysicalPointToIndex( testimg, truPoints ) )
predIndex = round( antsTransformPhysicalPointToIndex( testimg, predPoints ) )
# make point images
truPointsImage = predPointsImage = testimg * 0
for ( j in 1:nrow( truPoints ) ) {
  truPointsImage[truIndex[j,1],truIndex[j,2]]=j
  predPointsImage[predIndex[j,1],predIndex[j,2]]=j
  }
plot( testimg, iMath( truPointsImage, "GD",2) )
plot( testimg, iMath( predPointsImage, "GD",2) )
```

## look at some of the underlying images

Visualize the predicted images.

```{r vizi,echo=TRUE,eval=FALSE}
layout(matrix(1:8,nrow=2))
wsub = 6
for ( wpt in 1:length(segregions) ) {
  preds = predict( unetLM, Xte[c(1,3:4)] )
  testimg = as.antsImage( Xte[[1]][wsub,,,1] )
  testmsk = as.antsImage( Xte[[3]][wsub,,,1] )
  testht = as.antsImage( Xte[[5]][wsub,,,wpt] )
  testcc = as.antsImage( Xte[[4]][wsub,,,2] ) # coord conv
  testhtp = as.antsImage( preds[[1]][wsub,,,wpt] )
  plot( testimg, testht )
  plot( testimg, testhtp )
  }
```

## Inference in completely out of sample data

Inference follows the same patterns as above but there are many ways that
one can implement it.  Please implement inference on your own.

**Run on CPU thread when done**

```{r inferenceB,echo=FALSE}
# basically the same as training and augmentation - please treat as an exercise
```

## Future work

The package will evolve with the associated needs of related projects.
NOTE: this example is really very trivial and is not meant to "work" on a
real problem.  However, it has been tested in very close form on real
non-trivial problems with very similar parameters.
